Multiway Spherical Clustering via \\ Degree-Corrected Tensor Block Models

### 2022-01-06 tensor_clustering/tensor_clustering_ieee/ieee.tex
update

Suppose we have an order-$K$ data tensor $\tY \in \bbR^{p \times \cdots \times p}$. For ease of notation, we focus on symmetric tensors in this section; our framework easily extends to general asymmetric tensors. Assume there exist $r \geq 2$  disjoint communities among the $p$ nodes. We represent the community assignment by a function $z \colon [p]\mapsto[r]$, where $z(i) = a$ for $i$-th node that belongs to the $a$-th community. Then, $z^{-1}(a)=\{i\in[p]\colon z(i)=a\}$ denotes the set of nodes that belong to the $a$-th community, and $|z^{-1}(a)|$ denotes the number of nodes in the $a$-th community.
%, and we let $p_a = \sum_{i \in [p]}\ind \offf{ z_i= a}$ denote the size of $a$-th community. 
Let $\mtheta=(\theta(1),\ldots,\theta(p))^T$ denote the degree heterogeneity for $p$ nodes. We consider the order-$K$ dTBM~\citep{ghoshdastidar2017consistency,ke2019community},
\begin{equation}\label{eq:model_margin}
    \tY(i_1,\ldots,i_K) = \tS(z(i_1),\ldots,z(i_K)) \prod_{k = 1}^K\theta_{i_k} + \tE(i_1,\ldots,i_K), 
\end{equation}
\normalsize
where $\tS \in \bbR^{r \times \cdots \times r}$ is an order-$K$ tensor collecting the block means among communities, and $\tE\in\mathbb{R}^{p\times \cdots \times p}$ is a noise tensor consisting of independent mean-zero sub-Gaussian entries with variance bounded by $\sigma^2$. The unknown parameters are $z$, $S$, and $\mtheta$. The dTBM can be equivalently written in a compact form of tensor-matrix product:
\begin{equation}\label{eq:model_tensor}
 \mathbb{E}\tY = \tS \times_1 \mTheta \mM \times_2 \cdots \times_K  \mTheta \mM,
\end{equation}
where $\mTheta = \text{diag}(\theta(1),...,\theta(p)) \in \bbR^{p \times p}$ is a diagonal matrix, $\mM \in \offf{0,1}^{p \times r}$ is the membership matrix associated with community assignment $z$ such that $\mM(i,j)=\ind\{z(i)=j\}$. By definition, each row of $\mM$ has one copy of 1's and 0's elsewhere. Note that the discrete nature of $\mM$ renders our model~\eqref{eq:model_tensor} more challenging than Tucker decomposition. %The Tucker rank of the signal tensor $\bbE \tY$ is bounded by, but not necessarily equal to, $(r,...,r)$~\citep{wang2019multiway}. 
We call a tensor $\tX$ an $r$-block tensor with degree $\mtheta$
%with the core tensor $\tS$ 
if $\tX$ admits dTBM~\eqref{eq:model_tensor}. Here, we give two special cases of dTBM.

@@

\begin{algorithm}[h!]
\caption*{\bf Algorithm: Multiway spherical clustering for degree-corrected tensor block model }
\vspace{.15cm}
\begin{algorithmic}[1]
\Algphase{Sub-algorithm 1: Weighted higher-order initialization}
\INPUT Observation $\tY \in \bbR^{p\times \cdots \times p}$, number of cluster $r$, relaxation factor $\eta > 1$ in $k$-means clustering.
\State Compute factor matrix $ \mU_{\text{pre}} = \text{SVD}_{r} (\text{Mat}(\tY))$ and the $(K-1)$-mode projection $
\tX_{\text{pre}} = \tY \times_1  \mU_{\text{pre}}^T \times_2 \cdots \times_{K-1} \mU_{\text{pre}}^T.$
\State Compute factor matrix $\hat \mU = \text{SVD}_{r}(\text{Mat}(\tX_{\text{pre}}))$ and denoised tensor
$\hat \tX = \tY \times_1 \hat \mU \hat \mU^T \times_2 \cdots \times_K \hat \mU \hat \mU^T.$
\State {Let $\hat \mX = \text{Mat}(\hat \tX)$ and $I_0=\{i \in [p]: \onormSize{}{\hat \mX_{i:}} = 0\}$. Set $\hat z(i)$ randomly in $[r]$ for $i \in I_0$.}
\State{For all $i\in I_0^c$, compute normalized rows
$\hat \mX_{i:}^s :=\onormSize{}{\hat \mX_{i:}}^{-1} \hat \mX_{i:}.$
}
\State {Solve the clustering $\hat z\colon [p]\to[r]$ and centroids $ (\hat \mx_j)_{j\in[r_k]}$ using weighted $k$-means, such that}
\begin{align}
    &\sum_{i \in I_0^c }  \onormSize{}{\hat \mX_{i:}}^2 \onormSize{}{\hat \mX_{i:}^s - \hat \mx_{\hat z(i)} }^2 
    \leq 
    \eta \min_{\substack{\bar \mx_j, j\in[r], \bar z(i),i\in I_0^{c}}} \sum_{i \in S^c } \onormSize{}{\hat \mX_{i:}}^2 \onormSize{}{ \hat \mX_{i:}^s -   \bar \mx_{\bar z(i)}}^2.
\end{align}
\OUTPUT Initial clustering $z^{(0)} \leftarrow \hat z$.

@@

\begin{rmk}[Comparison to previous results] For fixed SNR, our initialization error rate with $K=2$ agrees with the initialization error rate $\tO(p)$ in matrix models~\citep{gao2018community}. Furthermore, in the special case of non-degree TBMs, we achieve the same initial misclassification error $\tO(p^{-K/2})$ as in non-degree models~\citep{han2020exact}. Theorem~\ref{thm:initial} implies the advantage of our algorithm in achieving both accuracy and model flexibility. 
%The result demonstrates the advantage of our algorithm in achieving both accuracy and model flexibility. 
\end{rmk}

@@

\begin{defn}[Locally linear stability] 
Define the $\varepsilon$-neighborhood of $z$ by $\tN(z,\epsilon)=\{\bar z\colon \ell(\bar z, z)\leq \epsilon\}$. We define two cluster-size vectors for $\bar z$,
\begin{align}
    &\mp(\bar z)=(|\bar z^{-1}(1)|, \ldots,|\bar z^{-1}(r)|)^T, \quad \mp_{\mtheta}(\bar z)=(\onormSize{}{\mtheta_{\bar z^{-1}(1)}}_1,\ldots,\onormSize{}{\mtheta_{\bar z^{-1}(r)}}_1)^T.
\end{align}

@@

{\bf Parameter Selection.} Note that we assume the number of clusters $r$ is given for the algorithm. We impose the Bayesian information criterion (BIC)
and choose the number of cluster that minimizes BIC, i.e., under dTBM~\eqref{eq:model_tensor},
\begin{equation}
    \hat r = \argmin_{r \in \mathbb{Z}_+ } \of{ -p^K \tL_{\tY}(\hat z(r), \hat \tS(r), \hat \mtheta(r)) + p_e(r) K \log p},
\end{equation}
where the triplet $(\hat z(r), \hat \tS(r), \hat \mtheta(r))$ are estimated parameters with given number of cluster $r$, $\tL_{\tY}$ denote the log-likelihood with observations $\tY$, and $p_e = r^K + p(\log r + 1)$ is the effective number of parameters. Particularly, we obtain the estimate $\hat \mtheta(r)$ by first calculating the reduced tensor $\tY^{\text{d}}$ with $\hat z(r)$ and then normalizing the row norms $\onormSize{}{\mY_{i:}^{\text{d}}}$ to 1 in each cluster. With Gaussian observation $\tY$, we have
\begin{equation}
    \tL_{\tY} = - \log(\onormSize{}{\hat \tX - \tY}_F^2), \quad \hat \tX = \tS(r) \times_1 \hat \mTheta(r) \hat \mM(r) \times_2 \cdots \times_K \hat \mTheta(r) \hat \mM(r).
\end{equation}
We evaluate the performance of the BIC criterion in Section~\ref{subsec:num_theory}.

@@

{
\color{red}

@@

\begin{enumerate}
\item \textbf{Angle-based techniques} 
\begin{itemize}
    \item Padding of the vectors won't extremely change the angle gap. 
    \item Using geometry property to upper bound $I_{11}, J_{11}$. (have not found a good name for this technique yet)
\end{itemize}
    \item \textbf{Proof idea of Theorem 4}
    \item \textbf{Proof idea of Theorem 5}
\end{enumerate}

@@

\paragraph{Notations.}
\begin{enumerate}
    \item For all $ k \in [K]$, denote the tensor matricizations as
    \begin{equation}
        \mY_k = \mat_k \of{ \tY }, \quad \mS_k = \mat_k \of{\tS}, \quad \mE_k = \mat_k \of{ \tE}, \quad \mX_k = \mat_k \of{\tX}.
    \end{equation}
    %We ignore the subscript $k$ when considering symmetric tensors. 
    \item For a vector $\ma$, let $\ma^{s} \coloneqq \ma/\onorm{\ma}$ denote the normalized vector. We make the convention that $\ma^s = 0$ if $\ma = 0$. 
    % Then we rewrite the angle gap~\eqref{eq:general_minial_gap} as
    % \begin{equation}
    %     \Delta_{\min} = \min_{k \in [K]} \min_{a \neq b \in [r_k]} \onorm{ \mS_{k,a:}^s - \mS_{k, b:}^s }.
    % \end{equation}
    \item For a matrix $\mA \in \bbR^{n \times m} \in \bbR^{n^K \times m^K}$, let $\mA^{\otimes K}$ denotes the kronecker product of $K$ matrices $\mA \otimes \cdots \otimes \mA$,
    \item For two terms $a$ and $b$, let $a \asymp b$ if there exist two positive constants $c, C$ such that $cb \leq a\leq Cb$. 
\end{enumerate}

 \paragraph{Generalized dTBM.}

@@

% \begin{equation}\label{eq:general_family}
%     \tP (r)= \offf{ \of{\{z_k\}, \tS, \{\mtheta_k\}}: \mtheta_k \in\mathbb{R}^p_{+}, {c_1p\over r}\leq |z_k^{-1}(a)| \leq {c_2 p\over r}, c_3\leq \onorm{\mat_k(\tS)}_{a:}\leq c_4 , \onormSize{}{\mtheta_{z_k^{-1}(a)}}_1=|z_k^{-1}(a)|, a \in [r], k\in[K]}.
% \end{equation}
\begin{align}
    \tP(\{r_k\}) = \Bigg\{ \of{\{z_k\}, \tS, \{\mtheta_k\}}:\  &\mtheta_k \in\mathbb{R}^p_{+}, {c_1 p_k\over r_k} |z_k^{-1}(a)| \leq {c_2 p_k\over r_k}, c_3 \leq  \onorm{\mS_{k,a:}} \leq c_4 ,\\
    &\onormSize{}{\mtheta_{k,z_k^{-1}(a)}}_1=|z_k^{-1}(a)|, \text{ for all } a \in [r_k], k\in[K]\Bigg\}. \label{eq:general_family}
\end{align}
% \begin{equation}
%     \tP = \offf{ \of{\{z_k\}, \tS, \{\mtheta_k\}}: \mtheta_k \in\mathbb{R}^p_{+}, |z_k^{-1}(a)| \asymp {p_k\over r_k},  \onorm{\mS_{k,a:}} \asymp 1 , \onormSize{}{\mtheta_{k,z_k^{-1}(a)}}_1=|z_k^{-1}(a)|, a \in [r], k\in[K]}.
% \end{equation}

@@

\paragraph{Useful Lemma for the Proof of Theorem~\ref{thm:unique}}

@@

\paragraph{Useful Definitions and Lemmas for the Proof of Theorem~\ref{thm:comp}} 
\begin{defn}[Principal angles] For two matrices $\mU, \hat \mU \in \mathbb{O}^{p \times r}$ %\fixme{Miaoyan}{do we need column organization  $\mathbb{O}(p,r)?$. Yes.}
, we define the principal angles between $\mU$ and $\hat \mU$ by the following diagonal matrix
\begin{equation}
    \text{Angle}(\mU, \hat \mU) = \text{diag}( \arccos(\sigma_1), \cdots, \arccos (\sigma_r) ),
\end{equation}
where $\sigma_1 \geq \cdots \geq \sigma_r>0$ are singular values of $\mU^T \hat \mU$. 
%\fixme{Miaoyan}{Geometry intuition. $\text{Angle}(\mU \mx, \hat \mU \mx) \leq \sigma_r$ for all $\mx$?} 
We define the norm of the principal angles as 
\begin{equation}
     \onorm{\sin(\mU, \hat \mU)}_{\sigma} = \max_{i \in [r]} \sin ( \arccos \of{ \sigma_i} ).
\end{equation}
%\fixme{Miayan}{Properties: $\onormSize{}{\sin(\mU, \hat \mU)}_\sigma= \onormSize{}{\mU^{\perp}\hat \mU}_{\sigma}$. $\max_{\mx\in \mathbb{R}^r}\sin \text{Angle}(\mU \mx, \hat \mU x)\leq \onormSize{}{\mU^{\perp}\hat \mU}_{\sigma}$}

@@

\begin{lem}[Angle distance and misclassification error]\label{lem:angle_mcr} Consider two membership matrices $\mM, \hat \mM \in \bbR^{p \times r}$ with corresponding assignments $z, \hat z: [p] \mapsto [r]$. If the angle distance $\onorm{\sin (\mM, \hat \mM)}_{\sigma} \geq c$ for a constant $c >0$, then the misclassification error $\min_{\pi \in \Pi}p\ell(\hat z,\pi \circ z )\geq 1$.
\end{lem}

@@

\section*{Proof of Theorem~\ref{thm:initial}}
\begin{proof}[Proof of Theorem 4] Recall that we prove the Theorem~\ref{thm:initial} under the dTBM~\eqref{eq:model_tensor} with parameters $(z, \tS, \mtheta)$. We drop the subscript $k$ in the matricizations $\mX_k, \mS_k, \mX_k$. For simplicity, let $\hat z$ denote the output of Sub-Algorithm~\hyperref[alg:main]{1} $\hat z^{(0)}$.

>>>>

Suppose we have an order-$K$ data tensor $\tY \in \bbR^{p \times \cdots \times p}$. For ease of notation, we focus on symmetric tensors in this section; our framework easily extends to general asymmetric tensors. Assume there exist $r \geq 2$  disjoint communities among the $p$ nodes. We represent the community assignment by a function $z \colon [p]\mapsto[r]$, where $z(i) = a$ for $i$-th node that belongs to the $a$-th community. Then, $z^{-1}(a)=\{i\in[p]\colon z(i)=a\}$ denotes the set of nodes that belong to the $a$-th community, and $|z^{-1}(a)|$ denotes the number of nodes in the $a$-th community.
%, and we let $p_a = \sum_{i \in [p]}\ind \offf{ z_i= a}$ denote the size of $a$-th community. 
Let $\mtheta=(\theta(1),\ldots,\theta(p))^T$ denote the degree heterogeneity for $p$ nodes. We consider the order-$K$ dTBM~\citep{ghoshdastidar2017consistency,ke2019community},
\begin{equation}\label{eq:model_margin}
    \tY(i_1,\ldots,i_K) = \tS(z(i_1),\ldots,z(i_K)) \prod_{k = 1}^K\theta_{i_k} + \tE(i_1,\ldots,i_K), 
\end{equation}
\normalsize
where $\tS \in \bbR^{r \times \cdots \times r}$ is an order-$K$ tensor collecting the block means among communities, and $\tE\in\mathbb{R}^{p\times \cdots \times p}$ is a noise tensor consisting of independent mean-zero sub-Gaussian entries with variance bounded by $\sigma^2$. The unknown parameters are $z$, $S$, and $\mtheta$. The dTBM can be equivalently written in a compact form of tensor-matrix product:
\begin{equation}\label{eq:model_tensor}
 \mathbb{E}\tY = \tS \times_1 \mTheta \mM \times_2 \cdots \times_K  \mTheta \mM,
\end{equation}
where $\mTheta = \text{diag}(\theta(1),...,\theta(p)) \in \bbR^{p \times p}$ is a diagonal matrix, $\mM \in \offf{0,1}^{p \times r}$ is the membership matrix associated with community assignment $z$ such that $\mM(i,j)=\ind\{z(i)=j\}$. By definition, each row of $\mM$ has one copy of 1's and 0's elsewhere. Note that the discrete nature of $\mM$ renders our model~\eqref{eq:model_tensor} more challenging than Tucker decomposition. %The Tucker rank of the signal tensor $\bbE \tY$ is bounded by, but not necessarily equal to, $(r,...,r)$~\citep{wang2019multiway}. 
We call a tensor $\tY$ an $r$-block tensor with degree $\mtheta$
%with the core tensor $\tS$ 
if $\tY$ admits dTBM~\eqref{eq:model_tensor}. Here, we give two special cases of dTBM.

@@

\begin{algorithm}[h!]
\caption*{\bf Algorithm: Multiway spherical clustering for degree-corrected tensor block model }
\vspace{.15cm}
\begin{algorithmic}[1]
\Algphase{Sub-algorithm 1: Weighted higher-order initialization}
\INPUT Observation $\tY \in \bbR^{p\times \cdots \times p}$, number of cluster $r$, relaxation factor $\eta > 1$ in $k$-means clustering.
\State Compute factor matrix $ \mU_{\text{pre}} = \text{SVD}_{r} (\text{Mat}(\tY))$ and the $(K-1)$-mode projection $
\tX_{\text{pre}} = \tY \times_1  \mU_{\text{pre}}^T \times_2 \cdots \times_{K-1} \mU_{\text{pre}}^T.$
\State Compute factor matrix $\hat \mU = \text{SVD}_{r}(\text{Mat}(\tX_{\text{pre}}))$ and denoised tensor
$\hat \tX = \tY \times_1 \hat \mU \hat \mU^T \times_2 \cdots \times_K \hat \mU \hat \mU^T.$
\State {Let $\hat \mX = \text{Mat}(\hat \tX)$ and $S_0=\{i \in [p]: \onormSize{}{\hat \mX_{i:}} = 0\}$. Set $\hat z(i)$ randomly in $[r]$ for $i \in S_0$.}
\State{For all $i\in S_0^c$, compute normalized rows
$\hat \mX_{i:}^s :=\onormSize{}{\hat \mX_{i:}}^{-1} \hat \mX_{i:}.$
}
\State {Solve the clustering $\hat z\colon [p]\to[r]$ and centroids $ (\hat \mx_j)_{j\in[r_k]}$ using weighted $k$-means, such that}
\begin{align}
    &\sum_{i \in S_0^c }  \onormSize{}{\hat \mX_{i:}}^2 \onormSize{}{\hat \mX_{i:}^s - \hat \mx_{\hat z(i)} }^2 
    \leq 
    \eta \min_{\substack{\bar \mx_j, j\in[r], \bar z(i),i\in S_0^{c}}} \sum_{i \in S^c } \onormSize{}{\hat \mX_{i:}}^2 \onormSize{}{ \hat \mX_{i:}^s -   \bar \mx_{\bar z(i)}}^2.
\end{align}
\OUTPUT Initial clustering $z^{(0)} \leftarrow \hat z$.

@@

\begin{rmk}[Comparison to previous results] For fixed SNR, our initialization error rate with $K=2$ agrees with the initialization error rate $\tO(p)$ in matrix models~\citep{gao2018community}. Furthermore, in the special case of non-degree TBMs, we achieve the same initial misclustering error $\tO(p^{-K/2})$ as in non-degree models~\citep{han2020exact}. Theorem~\ref{thm:initial} implies the advantage of our algorithm in achieving both accuracy and model flexibility. 
%The result demonstrates the advantage of our algorithm in achieving both accuracy and model flexibility. 
\end{rmk}

@@

\begin{defn}[Locally linear stability] \label{def:stable}
Define the $\varepsilon$-neighborhood of $z$ by $\tN(z,\epsilon)=\{\bar z\colon \ell(\bar z, z)\leq \epsilon\}$. We define two cluster-size vectors for $\bar z$,
\begin{align}
    &\mp(\bar z)=(|\bar z^{-1}(1)|, \ldots,|\bar z^{-1}(r)|)^T, \quad \mp_{\mtheta}(\bar z)=(\onormSize{}{\mtheta_{\bar z^{-1}(1)}}_1,\ldots,\onormSize{}{\mtheta_{\bar z^{-1}(r)}}_1)^T.
\end{align}

@@

{\bf Parameter Selection.} Note that we assume the number of clusters $r$ is given for the algorithm. We impose the Bayesian information criterion (BIC)
and choose the number of cluster that minimizes BIC, i.e., under dTBM~\eqref{eq:model_tensor},
\begin{equation}
    \hat r = \argmin_{r \in \mathbb{Z}_+ } \of{ -p^K \tL_{\tY}(\hat z(r), \hat \tS(r), \hat \mtheta(r)) + p_e(r) K \log p},
\end{equation}
where the triplet $(\hat z(r), \hat \tS(r), \hat \mtheta(r))$ are estimated parameters with given number of cluster $r$, $\tL_{\tY}$ denote the log-likelihood with observations $\tY$, and $p_e = r^K + p(\log r + 1) - 1$ is the effective number of parameters. Particularly, we obtain the estimate $\hat \mtheta(r)$ by first calculating the reduced tensor $\tY^{\text{d}}$ with $\hat z(r)$ and then normalizing the row norms $\onormSize{}{\mY_{i:}^{\text{d}}}$ to 1 in each cluster; i.e., 
\begin{equation}
    \hat \theta(i,r) = \frac{\onormSize{}{\hat \mY^{\text{d}}(r)_{i:}}}{\sum_{j: \hat z(j,r) =  \hat z(i,r)} \onormSize{}{\hat \mY^{\text{d}}(r)_{j:}}},
\end{equation}
where $\hat \mY^{\text{d}}(r) = \mat(\hat \tY^{\text{d}}(r))$ and 
\begin{equation}
    \hat \tY^{\text{d}}(r)(i,a_2,\ldots,a_K) = 
\text{Ave}\{\tY(i,i_2,\ldots,i_K): \hat z(i_k, r) = a_k, k \neq 1 \}.
\end{equation}

With Gaussian observation $\tY$, we have
\begin{equation}
    \tL_{\tY} = - \log(\onormSize{}{\hat \tX - \tY}_F^2), \quad \hat \tX = \tS(r) \times_1 \hat \mTheta(r) \hat \mM(r) \times_2 \cdots \times_K \hat \mTheta(r) \hat \mM(r).
\end{equation}
We evaluate the performance of the BIC criterion in Section~\ref{subsec:num_theory}.

@@

@@

In this section, we provide the proof sketches for Theorem~\ref{thm:initial} and Theorem~\ref{thm:refinement}. Detail proofs are provided in Appendix.

\subsection{Proof sketch of Theorem~\ref{thm:initial}}
The proof of Theorem~\ref{thm:initial} is mainly inspired by the proof idea of \citet[Lemma 1]{gao2018community}, and extra difficulties due to the angle gap assumption and tensor property are addressed in our proof. Specifically, we control the misclustering error by the estimation error of $\hat \tX$ calculated in Step 2.  We prove the following inequality
\begin{equation}\label{eq:proof_4}
    \ell(z^{(0)},z) \lesssim \frac{1}{p}\min_{\pi \in \Pi} \sum_{i: z^{(0)}(i) \neq \pi(z(i))} \theta(i)^2 \lesssim \frac{\sigma^2 r^{K-1}}{ \Delta_{\min}^2 p^K} \onormSize{}{\hat \tX - \tX}_F^2 \lesssim \frac{r^K p^{-K/2}}{\text{SNR}},
\end{equation}
where $\tX = \bbE[\tY]$ is the true mean. The first inequality in~\eqref{eq:proof_4} holds with the assumption $\min_{i \in [p]} \theta(i) \geq c$ in Theorem~\ref{thm:initial}. The second inequality relies on the intermediate conclusion that the angle gap of mean tensor $\tX$ is lower bounded by a function of core tensor minimal angle gap $\Delta_{\min}$ defined in Assumption~\ref{assmp:min_gap}. Let $\ma^{s} \coloneqq \ma/\onorm{\ma}$ denote the normalized vector and make the convention that $\ma^s = 0$ if $\ma = 0$. We show that
\begin{equation}\label{eq:proof_4_gap}
\min_{z(i) \neq z(j)} \onormSize{}{[\mX_{i:}]^s - [\mX_{j:}]^s} \gtrsim \Delta_{\min},
\end{equation}
where $\mX = \mat(\tX)$. The most challenging part in the proof of Theorem~\ref{thm:initial} lies in the derivation of inequality~\eqref{eq:proof_4_gap}, in which the proof of \cite{gao2018community} is no longer applicable due to different angle gap assumption in our dTBM. Extra padding technique in Lemma~\ref{lem:pad} and balance assumption~\eqref{eq:degree} are equipped to derive~\eqref{eq:proof_4_gap}. Last, we finish the proof of Theorem~\ref{thm:initial} by showing the third inequality with \citet[Proposition 1]{han2020exact}.

\subsection{Proof sketch of Theorem 5}
The proof of Theorem~\ref{thm:refinement} is mainly inspired by the proof idea of  \citet[Theorem 2]{han2020exact}, and extra polar coordinate based techniques are imposed due to the nuisance degree heterogeneity and angle gap assumption. We conduct the contraction property of the misclustering loss
\begin{equation}
     L^{(t)} = \frac{1}{p}  \sum_{i \in [p]} \theta(i) \sum_{b \in [r]}  \ind \offf{ z^{(t)}(i) = b } \onorm{ \off{ \mS_{ z(i):}  }^s - \off{ \mS_{b:}  }^s  }^2,
\end{equation}
and introduce the oracle estimators for core tensor and assignment
\begin{equation}
    \tilde \tS = \tY \times_1 \mW^T \times_2 \cdots \times_K \mW^T, 
    %\quad \tilde z(i) = \argmin_{a \in [r]} \onormSize{}{ { [\mat(\tY)_{i:} \mV]^s - [\mat(\tilde \tS)_{a:}]^s } }^2
\end{equation}
where $\mW = \mM \of{ \text{diag}(\ind_{p}^T \mM) }^{-1}$ is the weighted true membership matrix. Let $ \mV = \mW^{\otimes (K-1)}$ denote the Kronecker product of $K-1$ $\mW$ matrices, and similarly define $\mW^{(t)}, \mV^{(t)}$ with $\mM^{(t)}$ corresponding to $z^{(t)}$. To evaluate $L^{(t+1)}$, we consider the event 
\begin{equation}\label{eq:proof_5_event}
    \ind \offf{ z^{(t+1)}(i) = b } = \ind \offf{       \onormSize{}{ [ \mY_{i:} \mV^{(t)}  ]^s - [\mS_{b:}^{(t)}]^s }^2 \leq \onormSize{}{ [ \mY_{i:} \mV^{(t)}  ]^s - [\mS_{z(i):}^{(t)}]^s }^2} \leq A_{ib} + B_{ib}
\end{equation}
where $\mY = \mat(\tY)$, $ \mS = \mat(\tS)$, $\mS^{(t)} = \mat(\tS^{(t)})$ and
\begin{align}
        A_{ib} &= \ind \offf{\ang{ \mE_{i:} \mV, \off{  \tilde \mS_{z(i):} }^s - \off{  \tilde \mS_{b:} }^s } \lesssim -  \onorm{ \off{ \mS_{z(i):}  }^s - \off{ \mS_{b:}  }^s  }^2 },\\
        B_{ib} &= \ind \offf{\onorm{ \off{ \mS_{z(i):}  }^s - \off{ \mS_{b:}  }^s  }^2 \lesssim F_{ib}^{(t)} + G_{ib}^{(t)} + H_{ib}^{(t)} }.
\end{align}
The terms $F_{ib}^{(t)}, G_{ib}^{(t)}, H_{ib}^{(t)}$ are controlled by $z^{(t)}, \tS^{(t)}$, and detail definitions are in \eqref{eq:f}, \eqref{eq:g}, \eqref{eq:h}. Note that the event $A_{ib}$ only involves the oracle estimator while all the terms related to the $t$-th iteration estimator are in $B_{ib}$. Thus, the inequality~\eqref{eq:proof_5_event} decomposes the misclustering loss in the $(t+1)$-th iteration by the oracle loss and the loss in $t$-th iteration. This decomposition leads to the separation of statistical error and computational error in the final upper bound of Theorem~\ref{thm:refinement}.

Specifically, we prove the contraction inequality
\begin{equation}\label{eq:proof_5_ineq}
    L^{(t+1)} \lesssim \xi + \rho L^{(t)}, \quad \xi = \frac{1}{p}  \sum_{i \in [p]} \theta(i) \sum_{b \in [r]}  A_{ib} \onorm{ \off{ \mS_{ z(i):}  }^s - \off{ \mS_{b:}  }^s  }^2,
\end{equation}
where $\rho \in (0,1)$ is the contraction parameter, and we call $\xi$ the oracle loss. Controlling the probability of event $B_{ib}$ and obtaining the $\rho L^{(t)}$ term in the right hand side of~\eqref{eq:proof_5_ineq} is the most tricky and challenging part in the proof of Theorem~\ref{thm:refinement}. Note that the true and estimated core tensors are involved as normalized rows such as $\mS_{a:}^s, \tilde \mS_{a:}^s, [\mS^{(t)}_{a:}]^s$. The Cartesian coordinate based analysis in \cite{han2020exact} is no longer applicable in our case. Instead, we use the polar coordinate based analysis and the geometry property of trigonometric functions to derive the high probability upper bounds for $F_{ib}^{(t)}, G_{ib}^{(t)}, H_{ib}^{(t)}$.

Further, by sub-Gaussian concentration, we prove the high probability upper bound for oracle loss
\begin{equation}\label{eq:proof_5_xi}
    \xi  \lesssim \exp\of{- \frac{p^{K-1}\text{SNR}}{r^{K-1}}}.
\end{equation}
Combining the decomposition~\eqref{eq:proof_5_ineq} and the oracle bound~\eqref{eq:proof_5_xi}, we finish the proof of Theorem~\ref{thm:refinement}.

@@

{\bf Notations.}
\begin{enumerate}
    \item For all $ k \in [K]$, denote the tensor matricizations as
    \begin{equation}
        \mY_k = \mat_k \of{ \tY }, \quad \mS_k = \mat_k \of{\tS}, \quad \mE_k = \mat_k \of{ \tE}, \quad \mX_k = \mat_k \of{\tX}.
    \end{equation}
    %We ignore the subscript $k$ when considering symmetric tensors. 
    \item For a vector $\ma$, let $\ma^{s} \coloneqq \ma/\onorm{\ma}$ denote the normalized vector. We make the convention that $\ma^s = 0$ if $\ma = 0$. 
    % Then we rewrite the angle gap~\eqref{eq:general_minial_gap} as
    % \begin{equation}
    %     \Delta_{\min} = \min_{k \in [K]} \min_{a \neq b \in [r_k]} \onorm{ \mS_{k,a:}^s - \mS_{k, b:}^s }.
    % \end{equation}
    \item For a matrix $\mA \in \bbR^{n \times m} \in \bbR^{n^K \times m^K}$, let $\mA^{\otimes K}$ denotes the kronecker product of $K$ matrices $\mA \otimes \cdots \otimes \mA$,
    \item For a matrix $\mA$, let $\onormSize{}{\mA}_\sigma$ denote the spectral norm of matrix $\mA$, which is equal to the maximal singular value of $\mA$; let $\lambda_k(\mA)$ denote the $k$-th largest singular value of $\mA$; let $\onormSize{}{\mA}_F$ denote the Frobenius norm of matrix $\mA$.
    \item For two terms $a$ and $b$, let $a \asymp b$ if there exist two positive constants $c, C$ such that $cb \leq a\leq Cb$. 
\end{enumerate}

{\bf Generalized dTBM.}

@@

% \begin{equation}\label{eq:general_family}
%     \tP (r)= \offf{ \of{\{z_k\}, \tS, \{\mtheta_k\}}: \mtheta_k \in\mathbb{R}^p_{+}, {c_1p\over r}\leq |z_k^{-1}(a)| \leq {c_2 p\over r}, c_3\leq \onorm{\mat_k(\tS)}_{a:}\leq c_4 , \onormSize{}{\mtheta_{z_k^{-1}(a)}}_1=|z_k^{-1}(a)|, a \in [r], k\in[K]}.
% \end{equation}
\footnotesize
\begin{align}
    \tP(\{r_k\}) = \Bigg\{ \of{\{z_k\}, \tS, \{\mtheta_k\}}:\  &\mtheta_k \in\mathbb{R}^p_{+}, {c_1 p_k\over r_k} |z_k^{-1}(a)| \leq {c_2 p_k\over r_k}, c_3 \leq  \onorm{\mS_{k,a:}} \leq c_4 ,\onormSize{}{\mtheta_{k,z_k^{-1}(a)}}_1=|z_k^{-1}(a)|, a \in [r_k], k\in[K]\Bigg\}. \label{eq:general_family}
\end{align}
\normalsize
% \begin{equation}
%     \tP = \offf{ \of{\{z_k\}, \tS, \{\mtheta_k\}}: \mtheta_k \in\mathbb{R}^p_{+}, |z_k^{-1}(a)| \asymp {p_k\over r_k},  \onorm{\mS_{k,a:}} \asymp 1 , \onormSize{}{\mtheta_{k,z_k^{-1}(a)}}_1=|z_k^{-1}(a)|, a \in [r], k\in[K]}.
% \end{equation}

@@

{\bf Useful Lemma for the Proof of Theorem~\ref{thm:unique}}

@@

{\bf Useful Definitions and Lemmas for the Proof of Theorem~\ref{thm:comp}} 
\begin{defn}[Principal angles] For two matrices $\mU, \hat \mU \in \mathbb{O}^{p \times r}$ %\fixme{Miaoyan}{do we need column organization  $\mathbb{O}(p,r)?$. Yes.}
, we define the principal angles between $\mU$ and $\hat \mU$ by the following diagonal matrix
\begin{equation}
    \text{Angle}(\mU, \hat \mU) = \text{diag}( \arccos(\sigma_1), \cdots, \arccos (\sigma_r) ),
\end{equation}
where $\sigma_1 \geq \cdots \geq \sigma_r>0$ are singular values of $\mU^T \hat \mU$. 
%\fixme{Miaoyan}{Geometry intuition. $\text{Angle}(\mU \mx, \hat \mU \mx) \leq \sigma_r$ for all $\mx$?} 
We define the norm of the principal angles as 
\begin{equation}
     \onorm{\sin(\mU, \hat \mU)}_{\sigma} = \max_{i \in [r]} \sin ( \arccos \of{ \sigma_i} ).
\end{equation}
%\fixme{Miayan}{Properties: $\onormSize{}{\sin(\mU, \hat \mU)}_\sigma= \onormSize{}{\mU^{\perp}\hat \mU}_{\sigma}$. $\max_{\mx\in \mathbb{R}^r}\sin \text{Angle}(\mU \mx, \hat \mU x)\leq \onormSize{}{\mU^{\perp}\hat \mU}_{\sigma}$}

@@

\begin{lem}[Angle distance and misclustering error]\label{lem:angle_mcr} Consider two membership matrices $\mM, \hat \mM \in \bbR^{p \times r}$ with corresponding assignments $z, \hat z: [p] \mapsto [r]$. If the angle distance $\onorm{\sin (\mM, \hat \mM)}_{\sigma} \geq c$ for a constant $c >0$, then the misclustering error $\min_{\pi \in \Pi}p\ell(\hat z,\pi \circ z )\geq 1$.
\end{lem}

@@

\newpage

\section*{Proof of Theorem~\ref{thm:initial}}
\begin{proof}[Proof of Theorem 4] We prove Theorem~\ref{thm:initial} under the dTBM~\eqref{eq:model_tensor} with parameters $(z, \tS, \mtheta)$. We drop the subscript $k$ in the matricizations $\mM_k, \mX_k, \mS_k$. For simplicity, let $\hat z$ denote the output of Sub-Algorithm~\hyperref[alg:main]{1}, $\hat z^{(0)}$.
